Inside Out: Externalizing Assumptions in Data Analysis as Validation Checks

### 2024-12-20 _manuscript/_tex/index.tex
update application results

The externalization of the data analysis process is a practice that has
potential to improve the trustworthiness of analyses in general
\citep{yu2024veridical} and the trustworthiness of subsequent products,
such as machine learning models, that may be built on such analyses.
While publication of analysis code and data is now a common a
requirement for the sake of reproducibility
\citep{peng2011reproducible}, the publication code alone is often
insufficient for understanding the thought process behind an analysis.
Code corresponding to published analyses often reflect the final
decisions made about analysis and do not reveal the decisions or
assumptions made about the data processing. Thus, a reader looking at
published code can often be left with many questions about why certain
choices were made. Developing an approach to reveal some of this process
without requiring a reader to essentially reconstruct the analysis
process from the raw data would provide an improved basis for trusting
an analysis result \citep{peng2021reproducible}.

@@

To quantify redundancy, the concept of mutual information is used.
Mutual information \(I(x, y)\) measures the amount of information shared
between two random variables and is defined as the KL-distance
\(D(p \parallel q)\) between the joint distribution of the two variables
and the product of the marginal distributions:

@@

\caption{\label{fig-logic-reg}Logic regression model fitted to the three
checks. The model suggests the rule (quantile(step, 0.6) \textgreater{}
10,000 OR quantile(step, 0.4) \textless{} 8,000) AND (NOT sd(step)
\textgreater{} 2,500). The NOT operator applied to
\texttt{sd(step)\ \textgreater{}\ 2,500)} is colored with a black
background to distinguish it from other checks.}

@@

Table~\ref{tbl-logic-reg} presents the calculated precision, recall, and
independence for the three individual checks and the check rule found by
the logic regression. The harmonic and arithmetic means are included to
combine the three measures. The results show that the two checks
produced by the logic regression can accurately predict 86.1\% cases of
all \emph{actual unexpected results} in the simulation data.
Furthermore, 43.1\% of all \emph{predicted unexpected results} were in
fact observed to be unexpected.

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\linewidth - 10\tabcolsep) * \real{0.5368}}
  >{\raggedleft\arraybackslash}p{(\linewidth - 10\tabcolsep) * \real{0.1053}}
  >{\raggedleft\arraybackslash}p{(\linewidth - 10\tabcolsep) * \real{0.0737}}
  >{\raggedleft\arraybackslash}p{(\linewidth - 10\tabcolsep) * \real{0.0737}}
  >{\raggedleft\arraybackslash}p{(\linewidth - 10\tabcolsep) * \real{0.0947}}
  >{\raggedleft\arraybackslash}p{(\linewidth - 10\tabcolsep) * \real{0.1158}}@{}}

@@

\tabularnewline

\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
checks
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
precision
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
recall
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
indep.
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
harmonic
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
arithmetic
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
check 1: q(step, 0.6) \textgreater{} 10000 & 0.319 & 0.575 & 1.000 &
0.511 & 0.631 \\
check 2: q(step, 0.4) \textless{} 8000 & 0.264 & 0.613 & 1.000 & 0.467 &
0.626 \\
check 3: sd(step) \textgreater{} 2500 & 0.153 & 0.289 & 1.000 & 0.273 &
0.481 \\
logic regression: (check 1 OR check 2) AND check 3 & 0.431 & 0.861 &
0.998 & 0.669 & 0.763 \\
regression tree & 0.542 & 0.780 & 1.000 & 0.727 & 0.774 \\

\end{longtable}

@@

This PM10 coefficient expectation can be framed as an analysis check
that fails, labelled as 1, if the estimate of the PM10 coefficient
(adjusted for temperature) is outside the range {[}0, 0.005{]}, and 0 if
it is within. Multiple factors can affect the estimated PM10
coefficient, such as the sample size, the strength of the correlation
between mortality and PM10, and the strength of the correlation between
mortality and temperature. Analysts may expect a reasonable sample size
to ensure the reliability of the coefficient estimate. Outliers in the
three variables can also leverage the coefficient. While these are
possible factors that could affect the analysis result, it is not clear
what the cutoff values for these checks should be to determine a
failure. Here we consider a list of checks in Table~\ref{tbl-checks}
with varied cutoff values for each:

\begin{longtable}[]{@{}l@{}}
\caption{A list of checks considered for the generalized linear model of
mortality on PM10 and temperature. The checks are based on the sample
size, correlation between mortality and PM10, correlation between
mortality and temperature, and univariate outlier detection. Multiple
cutoff values are specified for each check to determine a
failure.}\label{tbl-checks}\tabularnewline
\toprule\noalign{}
The check fails if \ldots{} \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
The check fails if \ldots{} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
Sample size less than or equal to 200 \\
Sample size less than or equal to 400 \\
Sample size less than or equal to 600 \\
Sample size less than or equal to 800 \\
Mortality-PM10 correlation less than \(-0.03\) \\
Mortality-PM10 correlation less than \(-0.04\) \\
Mortality-PM10 correlation less than \(-0.05\) \\
Mortality-PM10 correlation less than \(-0.06\) \\
Mortality-temperature correlation greater than \(-0.3\) \\
Mortality-temperature correlation greater than \(-0.35\) \\
Mortality-temperature correlation greater than \(-0.4\) \\
Mortality-temperature correlation greater than \(-0.45\) \\
Outlier(s) are presented in the variable PM10 \\
Outlier(s) are presented in the variable mortality \\
\end{longtable}

@@

A logic regression is fitted using all variations of the checks in
Table~\ref{tbl-checks} to predict whether the PM10 coefficient is
unexpected. Figure~\ref{fig-linear-reg-tree} shows the optimal logic
regression tree from the fitted model. Precision, recall, and
independence score, along with their harmonic and arithmetic mean are
calculated in Table~\ref{tbl-linear-reg}.

@@

\caption{\label{fig-linear-reg-tree}Logic regression model fitted to the
fourteen checks and the outcome expectation (unexpected) as the response
variable. The model suggests the relationship: (NOT sample size less
than or equal to 200 AND mortality-PM10 correlation \textless{} -0.04)
AND (mortality-temperature correlation \textgreater{} -0.35 OR there
exist mortality outlier) to predict the unexpected PM10 coefficient. The
NOT operator applied to \texttt{smpl\_size\ \textless{}=\ 200} is
colored with a black background to distinguish it from other checks.}

@@

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\linewidth - 10\tabcolsep) * \real{0.6106}}
  >{\raggedleft\arraybackslash}p{(\linewidth - 10\tabcolsep) * \real{0.0885}}
  >{\raggedleft\arraybackslash}p{(\linewidth - 10\tabcolsep) * \real{0.0619}}
  >{\raggedleft\arraybackslash}p{(\linewidth - 10\tabcolsep) * \real{0.0619}}
  >{\raggedleft\arraybackslash}p{(\linewidth - 10\tabcolsep) * \real{0.0796}}
  >{\raggedleft\arraybackslash}p{(\linewidth - 10\tabcolsep) * \real{0.0973}}@{}}

\caption{\label{tbl-linear-reg}Accuracy (precision and recall) and
parsimony (independence) metrics derived from the logic regression
model, along with harmonic and arithmetic means, for individual checks
(1: sample size, 2: mortality-PM10 correlation, 3: mortality-temperature
correlation, 4: mortality outlier), and the combined check rule 5:
(sample size AND mortality-PM10 correlation) AND (mortality-temperature
correlation OR mortality outlier.}

\tabularnewline

\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
check
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
precision
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
recall
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
indep.
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
harmonic
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
arithmetic
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
check 1: sample size \textless= 200 & 0.087 & 0.215 & 1 & 0.175 &
0.434 \\
check 2: cor(m, PM10) \textless{} -0.03 & 0.988 & 0.610 & 1 & 0.822 &
0.866 \\
check 3: cor(m, tmp) \textgreater{} -0.35 & 0.392 & 0.581 & 1 & 0.569 &
0.658 \\
check 4: mortality\_outlier & 0.649 & 0.641 & 1 & 0.732 & 0.763 \\
logic regression: (not check 1 AND check 2) AND (check 3 OR check 4) &
0.760 & 0.880 & 1 & 0.869 & 0.880 \\

\end{longtable}

@@

\begin{itemize}
\tightlist
\item
  sample size \emph{larger than} 200
\item
  mortality-temperature correlation greater than \(-0.35\)
\item
  mortality-PM10 correlation less than \(-0.04\)
\item
  mortality data contains outliers that are detected by the univariate
  outlier detection
\end{itemize}

The tree structure suggests checking mortality-PM10 correlation and a
sample size larger than 200 with an additional check of either outlier
on mortality or correlation between mortality and temperature. This
combined check rule generates a precision of 0.76 and a recall of 0.88
for predicting the unexpected PM10 coefficient. The single check,
\texttt{cor(m,\ PM10)\ \textless{}\ -0.03}, is also powerful with a high
precision of 0.988, but the low recall value of 0.61 suggests its high
false positive rate, as compared to the combined rule suggested by the
logic regression.

@@

In order to test our development of analysis validation checks based on
the New York City data, we incorporate data from 39 additional cities
from the original NMMAPS study and apply our checks to those data. These
cities (along with New York City) represent approximately the largest 40
metropolitan areas in the United States. The complete list of cities
used are presented in Table~\ref{tbl-all-cities}. For each city, we had
daily data on all-cause mortality, PM10, and average temperature for the
years 1992--2000.

@@

\caption{\label{tbl-all-cities}}

@@

\centering
\resizebox{\ifdim\width>\linewidth\linewidth\else\width\fi}{!}{
\begin{tabular}{lrccrrcc}
\toprule
City & \makecell[l]{PM10\\Coefficient} & \makecell[l]{Outcome\\Unexpected} & \makecell[l]{Predicted\\Unexpected} & \makecell[l]{Correlation\\Mortality\\PM10} & \makecell[l]{Correlation\\Mortality\\Temperature} & \makecell[l]{Mortality\\Outlier} & \makecell[l]{PM10\\Outlier}\\
\midrule
Atlanta & -0.0008 & T & T & --0.26 (T) & --0.16 (T) & F & F\\
Austin & -0.0030 & T & T & --0.12 (T) & --0.11 (T) & F & T\\
Baltimore & 0.0018 & F & F & --0.21 (T) & 0.01 (F) & T & T\\
Boston & 0.0019 & F & F & --0.16 (T) & 0.02 (F) & F & T\\
Buffalo & 0.0028 & F & F & --0.23 (T) & --0.02 (F) & F & T\\
\addlinespace
Chicago & 0.0011 & F & F & --0.38 (F) & --0.02 (F) & T & T\\
Cincinnati & 0.0011 & F & T & --0.27 (T) & --0.06 (T) & T & T\\
Cleveland & 0.0010 & F & F & --0.30 (F) & --0.02 (F) & T & T\\
Columbus & 0.0004 & F & T & --0.20 (T) & --0.07 (T) & F & T\\
Denver & 0.0010 & F & F & --0.23 (T) & 0.06 (F) & F & T\\
\addlinespace
Detroit & 0.0011 & F & F & --0.30 (T) & 0.04 (F) & F & T\\
Dallas/Fort Worth & 0.0016 & F & T & --0.30 (F) & --0.03 (T) & T & T\\
Honolulu & 0.0006 & F & F & --0.17 (T) & 0.05 (F) & T & T\\
Houston & 0.0007 & F & F & --0.22 (T) & --0.03 (F) & F & T\\
Indianapolis & 0.0014 & F & F & --0.13 (T) & 0.02 (F) & F & T\\
\addlinespace
Kansas City & 0.0022 & F & F & --0.20 (T) & 0.02 (F) & F & T\\
Los Angeles & 0.0011 & F & F & --0.47 (F) & --0.01 (F) & T & T\\
Las Vegas & 0.0008 & F & F & --0.27 (T) & 0.05 (F) & T & T\\
Memphis & -0.0007 & T & T & --0.16 (T) & --0.10 (T) & F & T\\
Miami & -0.0004 & T & F & --0.16 (T) & --0.02 (F) & F & T\\
\addlinespace
Milwaukee & 0.0026 & F & F & --0.22 (T) & 0.09 (F) & F & T\\
Minneapolis/St. Paul & 0.0009 & F & F & --0.30 (F) & --0.01 (F) & T & T\\
New York & 0.0022 & F & F & --0.46 (F) & --0.01 (F) & T & T\\
Oakland & 0.0008 & F & F & --0.28 (T) & 0.04 (F) & F & T\\
Orlando & -0.0008 & T & T & --0.21 (T) & --0.05 (T) & F & T\\
\addlinespace
Philadelphia & 0.0021 & F & F & --0.27 (T) & 0.11 (F) & T & T\\
Phoenix & 0.0009 & F & F & --0.44 (F) & 0.07 (F) & T & T\\
Pittsburgh & 0.0009 & F & T & --0.36 (F) & --0.08 (T) & T & T\\
Riverside & 0.0001 & F & T & --0.27 (T) & --0.11 (T) & T & T\\
Sacramento & 0.0005 & F & F & --0.30 (T) & 0.05 (F) & T & T\\
\addlinespace
Salt Lake City & -0.0004 & T & T & --0.19 (T) & --0.03 (T) & F & T\\
San Antonio & -0.0021 & T & T & --0.22 (T) & --0.16 (T) & F & T\\
San Bernardino & 0.0006 & F & T & --0.28 (T) & --0.09 (T) & T & T\\
San Diego & 0.0011 & F & F & --0.33 (F) & 0.03 (F) & T & T\\
San Jose & 0.0006 & F & F & --0.23 (T) & 0.08 (F) & F & T\\
\addlinespace
Seattle & 0.0003 & F & F & --0.24 (T) & 0.06 (F) & F & T\\
Santa Ana/Anaheim & 0.0005 & F & T & --0.26 (T) & --0.04 (T) & T & T\\
St. Petersburg & 0.0012 & F & F & --0.31 (F) & 0.01 (F) & F & T\\
Tampa & -0.0003 & T & T & --0.20 (T) & --0.04 (T) & F & T\\
Tucson & 0.0028 & F & F & --0.31 (F) & 0.18 (F) & T & T\\
\bottomrule
\end{tabular}}

@@

We applied each of the analysis validation checks and the fitted logic
regression model separately to each of the cities' data to make
predictions of whether the outcome will be unexpected or not. We also
applied the generalized linear model to the data from each of the cities
to estimate the association between PM10 and mortality in each city. The
estimates of the association and the results of each of the analysis
validation checks for each city are shown in Table~\ref{tbl-all-cities}.

@@

\caption{\label{tbl-accuracy}}

@@

\caption{A 2x2 table of the outcome and the predicted unexpected results from the 40 cities.}
\centering
\begin{tabular}[t]{lrr}
\toprule
\multicolumn{1}{c}{} & \multicolumn{2}{c}{Prediction} \\
\cmidrule(l{3pt}r{3pt}){2-3}
Actual & FALSE & TRUE\\
\midrule
FALSE & 25 & 7\\
TRUE & 1 & 7\\
\bottomrule
\end{tabular}

@@

Table~\ref{tbl-accuracy} summaries the prediction accuracy of the logic
regression model for the 40 cities. The model correctly predicts
unexpected outcomes in 32 cities, producing an accuracy of 80\%. Among
the 8 cities with an unexpected outcome (PM10 coefficient outside {[}0,
0.005{]}), the model correctly identifies 7 (Atlanta, Austin, Memphis,
Orlando, Salt Lake City, San Antonio, Tampa), giving a recall of 88\%.
These cities are flagged due to failure in both mortality-PM10
correlation and mortality-temperature correlation checks. Out of the 14
cities with positive predictions from the model, 7 cities (Cincinnati,
Columbus, Dallas/Fort Worth, Pittsburgh, Riverside, San Bernardino,
Santa Ana/Anaheim) were a false positive, resulting in a precision of
50\%.

{[}Sherry: Summarize the results from the 40 cities, along with 2x2
table of outcome vs.~predicted{]}

@@

@@

Systematically generating realistic simulated data is a key component of
our approach and is deserving of further consideration. In the PM10
example, the inverse transform method is used to preserve the
correlation structure among mortality, PM10, and temperature. However,
simulation can become complex when additional restrictions are imposed
or when a greater range of scenarios is desired. In such cases,
techniques like the acceptance-reject method or permutation may be used
to generate the data.

>>>>

The externalization of the data analysis process is a practice that has
potential to improve the trustworthiness of analyses in general
\citep{yu2024veridical} and the trustworthiness of subsequent products,
such as machine learning models, that may be built on such analyses.
While publication of analysis code and data is now a common a
requirement for the sake of reproducibility
\citep{peng2011reproducible}, the published code alone is often
insufficient for understanding the thought process behind an analysis.
Code corresponding to published analyses often reflect the final
decisions made about analysis and do not reveal the decisions or
assumptions made about the data processing. Thus, a reader looking at
published code can often be left with many questions about why certain
choices were made. Developing an approach to reveal some of this process
without requiring a reader to essentially reconstruct the analysis
process from the raw data would provide an improved basis for trusting
an analysis result \citep{peng2021reproducible}.

@@

To quantify redundancy, the concept of mutual information is used.
Mutual information \(I(x, y)\) measures the amount of information shared
between two random variables and is defined as the Kullback-Leibler
distance \(D(p \parallel q)\) between the joint distribution of the two
variables and the product of the marginal distributions:

@@

\caption{\label{fig-logic-reg}Logic regression model fitted to the three
checks. The model suggests the rule (quantile(step, 0.6) \textgreater{}
10,000 OR quantile(step, 0.4) \textless{} 8,000) AND (NOT sd(step)
\textgreater{} 2,500). The NOT operator applied to
\texttt{sd(step)\ \textgreater{}\ 2,500} is colored with a black
background to distinguish it from other checks.}

@@

Table~\ref{tbl-logic-reg} presents the calculated precision, recall, and
independence for the three individual checks and the check rule found by
the logic regression. The harmonic and arithmetic means are included to
combine the three measures. The results show that the three checks
produced by the logic regression can accurately predict 86.1\% cases of
all \emph{actual unexpected results} in the simulation data.
Furthermore, 43.1\% of all \emph{predicted unexpected results} were in
fact observed to be unexpected.

@@

\centering
\resizebox{\ifdim\width>\linewidth\linewidth\else\width\fi}{!}{
\begin{tabular}{>{\raggedright\arraybackslash}p{17em}rrrrr}
\toprule
checks & precision & recall & indep. & harmonic & arithmetic\\
\midrule
Check 1: q(step, 0.6) > 10000 & 0.319 & 0.575 & 1.000 & 0.511 & 0.631\\
Check 2: q(step, 0.4) < 8000 & 0.264 & 0.613 & 1.000 & 0.467 & 0.626\\
Check 3: sd(step) > 2500 & 0.153 & 0.289 & 1.000 & 0.273 & 0.481\\
Logic regression: (check 1 OR check 2) AND (not check 3) & 0.861 & 0.431 & 0.998 & 0.669 & 0.763\\
Regression tree & 0.542 & 0.780 & 1.000 & 0.727 & 0.774\\
\bottomrule
\end{tabular}}

@@

This PM10 coefficient expectation can be framed as an analysis check
that fails, labelled as 1, if the estimate of the PM10 coefficient
(adjusted for temperature) is outside the range {[}0, 0.005{]}, and 0 if
it is within. Multiple factors can affect the estimated PM10
coefficient, such as the strength of the correlation between mortality
and PM10, or between mortality and temperature. Outliers in the three
variables can also leverage the coefficient. While these are possible
factors that could affect the analysis result, it is not clear what the
cutoff values for these checks should be to determine a failure. Here we
consider a list of checks in Table~\ref{tbl-checks} with varied cutoff
values for each:

\begin{longtable}[]{@{}l@{}}
\caption{A list of checks considered for the generalized linear model of
mortality on PM10 and temperature. The checks are based on the sample
size, correlation between mortality and PM10, correlation between
mortality and temperature, and univariate outlier detection. Multiple
cutoff values are specified for each check to determine a
failure.}\label{tbl-checks}\tabularnewline
\toprule\noalign{}
The check fails (encoded as 1) if \ldots{} \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
The check fails (encoded as 1) if \ldots{} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
Mortality-PM10 correlation less than \(-0.05\) \\
Mortality-PM10 correlation less than \(-0.03\) \\
Mortality-PM10 correlation greater than \(0.03\) \\
Mortality-PM10 correlation greater than \(0.05\) \\
Mortality-temperature correlation greater than \(-0.3\) \\
Mortality-temperature correlation greater than \(-0.35\) \\
Mortality-temperature correlation greater than \(-0.4\) \\
Mortality-temperature correlation greater than \(-0.45\) \\
Outlier(s) are presented in the variable PM10 \\
Outlier(s) are presented in the variable mortality \\
\end{longtable}

@@

A logic regression is fitted using all variations of the checks in
Table~\ref{tbl-checks} to predict whether the PM10 coefficient is
unexpected. Given the imbalance of outcome (expected vs.~unexpected),
inverse weights proportional to the number of observations in each
outcome are applied during the logic regression fit.
Figure~\ref{fig-linear-reg-tree} shows the optimal logic regression tree
from the fitted model. Precision, recall, and independence score, along
with their harmonic and arithmetic mean are calculated in
Table~\ref{tbl-linear-reg}.

@@

\caption{\label{fig-linear-reg-tree}Logic regression model fitted to the
fourteen checks and the outcome expectation (unexpected) as the response
variable. The model suggests the relationship: (mortality-PM10
correlation \textless{} -0.03) AND (mortality-temperature correlation
\textgreater{} -0.3 OR (there exist mortality outlier AND there exist
PM10 outlier)) to predict the unexpected PM10 coefficient.}

@@

\caption{\label{tbl-linear-reg}Accuracy (precision and recall) and
parsimony (independence) metrics derived from the logic regression model
(last row) and individual checks, along with harmonic and arithmetic
means.}

\centering
\resizebox{\ifdim\width>\linewidth\linewidth\else\width\fi}{!}{
\begin{tabular}{>{\raggedright\arraybackslash}p{18em}rrrrr}
\toprule
check & precision & recall & indep. & harmonic & arithmetic\\
\midrule
Check 1: cor(m, PM10) < -0.03 & 0.551 & 0.894 & 1 & 0.763 & 0.815\\
Check 2: cor(m, tmp) > -0.35 & 0.281 & 0.556 & 1 & 0.472 & 0.612\\
Check 3: mortality outlier & 0.302 & 0.775 & 1 & 0.536 & 0.692\\
Check 4: PM10 outlier & 0.305 & 0.766 & 1 & 0.537 & 0.690\\
Logic regression: (check 1) AND ((check 2) OR (check 3 AND check 4)) & 0.722 & 0.876 & 1 & 0.851 & 0.866\\
\bottomrule
\end{tabular}}

@@

\begin{itemize}
\tightlist
\item
  mortality-PM10 correlation less than \(-0.03\)
\item
  mortality-temperature correlation greater than \(-0.3\)
\item
  PM10 data contains outliers that are detected by the univariate
  outlier detection
\item
  mortality data contains outliers that are detected by the univariate
  outlier detection
\end{itemize}

The tree structure suggests checking mortality-PM10 correlation and a
sample size larger than 200 with an additional check of either outlier
on mortality or correlation between mortality and temperature. This
combined check rule generates a precision of 0.722 and a recall of 0.876
for predicting the unexpected PM10 coefficient.

@@

In order to test our development of analysis validation checks based on
the New York City data, we incorporate data from 39 additional cities
from the original NMMAPS study and apply our checks to those data. These
cities (along with New York City) represent approximately the largest 40
metropolitan areas in the United States by population. The complete list
of cities used are presented in Table~\ref{tbl-all-cities}. For each
city, we had daily data on all-cause mortality, PM10, and average
temperature for the years 1992--2000.

@@

\caption{\label{tbl-all-cities}Results from full analysis of 40 cities
from the NMMAPS air pollution and mortality study. In the table, a ``T''
indicates that a check failed and an ``F'' indicates that a check
passed.}

@@

\centering
\resizebox{\ifdim\width>\linewidth\linewidth\else\width\fi}{!}{
\begin{tabular}{lrccrrcc}
\toprule
City & \makecell[l]{PM10\\Estimated\\Coefficient} & \makecell[l]{Outcome\\Observed\\Unexpected} & \makecell[l]{Logic Regression\\Predicted\\Unexpected} & \makecell[l]{Correlation:\\Mortality-\\Temperature} & \makecell[l]{Correlation:\\Mortality-\\PM10} & \makecell[l]{Mortality\\Outlier} & \makecell[l]{PM10\\Outlier}\\
\midrule
Atlanta & -0.0008 & T & T & --0.26 (T) & --0.16 (T) & F & F\\
Austin & -0.0030 & T & T & --0.12 (T) & --0.11 (T) & F & T\\
Baltimore & 0.0018 & F & F & --0.21 (T) & 0.01 (F) & T & T\\
Boston & 0.0019 & F & F & --0.16 (T) & 0.02 (F) & F & T\\
Buffalo & 0.0028 & F & F & --0.23 (T) & --0.02 (F) & F & T\\
\addlinespace
Chicago & 0.0011 & F & F & --0.38 (F) & --0.02 (F) & T & T\\
Cincinnati & 0.0011 & F & T & --0.27 (T) & --0.06 (T) & T & T\\
Cleveland & 0.0010 & F & F & --0.30 (F) & --0.02 (F) & T & T\\
Columbus & 0.0004 & F & T & --0.20 (T) & --0.07 (T) & F & T\\
Denver & 0.0010 & F & F & --0.23 (T) & 0.06 (F) & F & T\\
\addlinespace
Detroit & 0.0011 & F & F & --0.30 (T) & 0.04 (F) & F & T\\
Dallas/Fort Worth & 0.0016 & F & T & --0.30 (F) & --0.03 (T) & T & T\\
Honolulu & 0.0006 & F & F & --0.17 (T) & 0.05 (F) & T & T\\
Houston & 0.0007 & F & F & --0.22 (T) & --0.03 (F) & F & T\\
Indianapolis & 0.0014 & F & F & --0.13 (T) & 0.02 (F) & F & T\\
\addlinespace
Kansas City & 0.0022 & F & F & --0.20 (T) & 0.02 (F) & F & T\\
Los Angeles & 0.0011 & F & F & --0.47 (F) & --0.01 (F) & T & T\\
Las Vegas & 0.0008 & F & F & --0.27 (T) & 0.05 (F) & T & T\\
Memphis & -0.0007 & T & T & --0.16 (T) & --0.10 (T) & F & T\\
Miami & -0.0004 & T & F & --0.16 (T) & --0.02 (F) & F & T\\
\addlinespace
Milwaukee & 0.0026 & F & F & --0.22 (T) & 0.09 (F) & F & T\\
Minneapolis/St. Paul & 0.0009 & F & F & --0.30 (F) & --0.01 (F) & T & T\\
New York & 0.0022 & F & F & --0.46 (F) & --0.01 (F) & T & T\\
Oakland & 0.0008 & F & F & --0.28 (T) & 0.04 (F) & F & T\\
Orlando & -0.0008 & T & T & --0.21 (T) & --0.05 (T) & F & T\\
\addlinespace
Philadelphia & 0.0021 & F & F & --0.27 (T) & 0.11 (F) & T & T\\
Phoenix & 0.0009 & F & F & --0.44 (F) & 0.07 (F) & T & T\\
Pittsburgh & 0.0009 & F & T & --0.36 (F) & --0.08 (T) & T & T\\
Riverside & 0.0001 & F & T & --0.27 (T) & --0.11 (T) & T & T\\
Sacramento & 0.0005 & F & F & --0.30 (T) & 0.05 (F) & T & T\\
\addlinespace
Salt Lake City & -0.0004 & T & T & --0.19 (T) & --0.03 (T) & F & T\\
San Antonio & -0.0021 & T & T & --0.22 (T) & --0.16 (T) & F & T\\
San Bernardino & 0.0006 & F & T & --0.28 (T) & --0.09 (T) & T & T\\
San Diego & 0.0011 & F & F & --0.33 (F) & 0.03 (F) & T & T\\
San Jose & 0.0006 & F & F & --0.23 (T) & 0.08 (F) & F & T\\
\addlinespace
Seattle & 0.0003 & F & F & --0.24 (T) & 0.06 (F) & F & T\\
Santa Ana/Anaheim & 0.0005 & F & T & --0.26 (T) & --0.04 (T) & T & T\\
St. Petersburg & 0.0012 & F & F & --0.31 (F) & 0.01 (F) & F & T\\
Tampa & -0.0003 & T & T & --0.20 (T) & --0.04 (T) & F & T\\
Tucson & 0.0028 & F & F & --0.31 (F) & 0.18 (F) & T & T\\
\bottomrule
\end{tabular}}

@@

We applied each of the analysis validation checks and the fitted logic
regression model separately to each of the cities' data to make
predictions of whether the outcome will be unexpected or not. We also
applied the generalized linear model to the data from each of the cities
to estimate the association between PM10 and mortality in each city,
adjusted for temperature. The estimates of the association and the
results of each of the analysis validation checks for each city are
shown in Table~\ref{tbl-all-cities}.

@@

\caption{\label{tbl-accuracy}Summary of the observed and the predicted
unexpected PM10 coefficient results from the 40 NMMAPS cities using the
logic regression model.}

@@

\begin{tabular}{lrr}
\toprule
\multicolumn{1}{c}{} & \multicolumn{2}{c}{Predicted} \\
\cmidrule(l{3pt}r{3pt}){2-3}
Observed & FALSE & TRUE\\
\midrule
FALSE & 25 & 7\\
TRUE & 1 & 7\\
\bottomrule
\end{tabular}

@@

It is clear from Table~\ref{tbl-all-cities} that some of the estimated
PM10 coefficients are outside the expected range of {[}0, 0.005{]}.
Eight of the 40 cities had negative estimated coefficients and were
therefore considered unexpected by our original criterion. (It is
perhaps worth noting that none of the unexpected outcomes was in the
positive direction.) Table~\ref{tbl-accuracy} summarizes the prediction
accuracy of the logic regression model for the 40 cities. The model
correctly predicts the PM10 coefficient status of 32 cities, producing
an accuracy rate of 80\%. Among the 8 cities with an unexpected outcome
(PM10 coefficient outside {[}0, 0.005{]}), the model correctly
identifies 7 (Atlanta, Austin, Memphis, Orlando, Salt Lake City, San
Antonio, Tampa), giving a recall of 88\% (only Miami was not properly
classified by the logic regression model). These cities were flagged due
to failures in both the mortality-PM10 correlation and
mortality-temperature correlation analysis checks. Out of the 14 cities
with positive predictions from the model, 7 cities (Cincinnati,
Columbus, Dallas/Fort Worth, Pittsburgh, Riverside, San Bernardino,
Santa Ana/Anaheim) were false positives, resulting in a precision of
only 50\%. This precision value is substantially lower than what was
estimated by the simulation procedure (see Table~\ref{tbl-linear-reg})
and suggests that the simulation does not adequately capture some
features of the data generation process.

@@

In Section~\ref{sec-pm10-mortality} we used the analysis checks in a
diagnostic manner to examine the eight cities whose PM10 coefficients
were unexpected. There, we found that seven of the cities had a
mortality-PM10 correlation that was more negative than expected while
also having a mortality-temperature correlation that was somewhat larger
(less negative) than expected. What this result implies for the broader
analysis depends on a number of factors, but the interplay between
mortality, PM10, and temperature may warrant further investigation
\citep[see e.g.][]{welty2005acute}. Note that simply because the PM10
coefficient estimates were unexpected by our criterion, we do not mean
to imply that they are ``wrong'' in any sense. Indeed, negative
estimated coefficients have been found in other studies of this nature
\citep{bell2004ozone}. Rather, in this type of analysis, it may be that
the interval for expected results needs to be revised. Further, it may
be necessary to incorporate other outputs from the analysis, such as
measures of uncertainty.

@@

Systematically generating realistic simulated data is a key component of
our approach and is deserving of further consideration. In the PM10
example in Section~\ref{sec-pm10-mortality}, the inverse transform
method was used to preserve the correlation structure among mortality,
PM10, and temperature. However, the simulation process can become
complex when additional restrictions are imposed or when a greater range
of scenarios is desired. In such cases, techniques like the
acceptance-reject method or permutation may be used to generate the
data. The results in Section~\ref{sec-pm10-mortality} on the broader
NMMAPS dataset suggest that our simulation procedure may have been
inadequate in reflecting the range of possible configurations that the
data could take. In particular, our split-data approach, using New York
City to guide the simulation and then applying the analysis checks to
other cities, may not have been ideal. It may be worth exploring some
recent work in data thinning \citep{neufeld2024data} and data fission
\citep{leiner2023data} to create training datasets that are more
representative of future data.
